---
title: "Microarray Data Stomach Antrum Epithelia, Lgr4 KO - Reanalysis with LIMMA"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, messages=FALSE}
rm(list=ls())
library(limma)
library(pheatmap)
library(readxl)
library(xtable)
library(reshape2)
```

# Introduction

This document describes preprocessing and QC of mouse experiments (stomach Antrum +/- Lgr4 KO) hybridized as dual channel to Agilent 014868 micro array. 

```{r,  results='hide', messages=FALSE}
# set up the experimental design
exp_design = as.data.frame(read_excel("ExpDesign_AllArrays.xls",sheet = 1), stringsAsFactors=F)
exp_design = subset(exp_design, Experiment == "Jonas-Lgr4")
colnames(exp_design) = make.names(colnames(exp_design))

rownames(exp_design)=as.character(exp_design$Scan.Name)
exp_design$dye_swap = ifelse(exp_design$Cy5 %in% c("control","Lgr4_5 WT"), T, F)
exp_design$FileName = paste(exp_design$Scan.Name, ".txt", sep="")
exp_design$Cy3 = paste(exp_design$Cy3_treatment,exp_design$Cy3_infection,sep="_")
exp_design$Cy5 = paste(exp_design$Cy5_treatment,exp_design$Cy5_infection,sep="_")

setwd("../Data/Raw/")
RG =read.maimages(exp_design$FileName,source="agilent.median", columns=list(G="gProcessedSignal",R="rProcessedSignal"))
setwd("..")

ed = exp_design

# fix outdated chip annotations
new_anno_file = "../Data/Agilent_14868_annotations_2017-06-27.Rdata"
load(new_anno_file)
old_anno = RG$genes
take_over_cols = colnames(old_anno)[!colnames(old_anno) %in% c("GeneName","Description","SystematicName")]
tmp = old_anno[,take_over_cols]
tmp$index=1:nrow(tmp)
tmp = merge(tmp, anno_tab_14868, by.x="ProbeName", by.y="ProbeID", all.x=T, sort=F)
new_col_order = c(take_over_cols, colnames(tmp)[!colnames(tmp) %in% take_over_cols])
new_anno = tmp[order(tmp$index),new_col_order]

RG$genes = new_anno
```


## Data overview

### Samples 

```{r,  results='asis', warning=FALSE}
sel_cols = c("Scan.ID","Chip.Barcode","dye_swap","Cy3_treatment","Cy5_treatment")
tmp = ed[,sel_cols]
print.xtable(xtable(tmp,display=rep("s",ncol(tmp)+1), align=paste("|",rep("l|", ncol(tmp)+1)), file="" , include.rownames=F), type="html")
```

## Raw data

```{r, expr,  fig.width=8, fig.height=8}
boxplot(data.frame(log2(RG$G)),main="Green", las=2, names = ed[colnames(RG$G),"Scan.ID"])
boxplot(data.frame(log2(RG$R)),main="Red", las=2, names = ed[colnames(RG$G),"Scan.ID"])

plotDensities(RG)
```

# Within-array normalization

Ratios of R/G intensities (i.e. M-values) are normalized using method **LOESS**.

```{r,  fig.width=8, fig.height=8}
MA <- normalizeWithinArrays(RG, method="loess")
MA.avg <- avereps(MA, ID=MA$genes$ProbeName)

limma::plotMA(MA,array=1,status=ifelse(MA$genes$ControlType==1,RG$genes$ProbeName,"gene"),cex=1, legend=F)
plotDensities(MA, main="Within-array normalized densities")

op <- par(mar = c(14,4,4,2) + 0.1)
boxplot(data.frame(MA$M),main="M", las=2, cex.names=1, names = ed[colnames(MA$M),"Scan.ID"])
boxplot(data.frame(MA$A),main="A", las=2, names = ed[colnames(RG$G),"Scan.ID"])
par(op)
```

# Correlation of samples based on M-values (R/G ratios) after LOESS

```{r, cor,  fig.width=8, fig.height=8}
exp_matrix = MA.avg$M * matrix(rep(ifelse(ed$dye_swap, -1, 1),each=nrow(MA.avg$M)),nrow(MA.avg$M),ncol(MA.avg$M))
dswp = exp_design[colnames(exp_matrix),]$dye_swap
cy3_label = as.character(exp_design[colnames(exp_matrix),]$Cy3)
cy5_label = as.character(exp_design[colnames(exp_matrix),]$Cy5)
#repl = exp_design[colnames(exp_matrix),]$replicate
#array_label = paste(ifelse(dswp, cy3_label, cy5_label), "_vs_",ifelse(dswp,cy5_label, cy3_label),"_",dswp,"_",exp_design[colnames(exp_matrix),]$Tissue,"_",exp_design[colnames(exp_matrix),]$Experiment ,sep="")
array_label = paste(ifelse(dswp, cy3_label, cy5_label), "_vs_",ifelse(dswp,cy5_label, cy3_label),"_",dswp,"_",exp_design[colnames(exp_matrix),]$Tissue,"_",exp_design[colnames(exp_matrix),]$Experiment,"_",exp_design[colnames(exp_matrix),]$Cy5_replicate ,sep="")
colnames(exp_matrix) = array_label
cc = cor(exp_matrix,method="spearman")
pheatmap(cc, main="Correlation heatmap")
```

```{r, echo=FALSE}
ed$Short = paste(ifelse(dswp, cy3_label, cy5_label), "_vs_",ifelse(dswp,cy5_label, cy3_label),"_",dswp,"_",ed$Scan.ID,sep="")
ed$Condition = paste(ed$Cy3,  sep="_")
```


<!-- # PCA -->

<!-- ```{r, PCA,  fig.width=8, fig.height=8} -->
<!-- ed2=ed -->
<!-- rownames(ed2) = array_label -->
<!-- ## Primary Component Analysis on normalized data after adjusting for patient effect -->
<!-- norm_exp = exp_matrix -->
<!-- NA_rows = apply(norm_exp,1,function(x) sum(is.na(x))) -->
<!-- pca = prcomp(t(norm_exp[NA_rows==0,])) -->
<!-- #pca = prcomp(t(normalized$E), na.action=na.omit) -->
<!-- plot(pca$x[,1],pca$x[,2],type="p", xlab="1st principal component",ylab="2nd principal component", main="PCA on normalized expression data", ylim=c(-200,200), xlim=c(-150,200), pch=20) -->
<!-- cp = palette(rainbow(8)) -->
<!-- text(pca$x[,1],pca$x[,2],labels=ed2[colnames(norm_exp),]$Short, col=cp[as.numeric(as.factor(ed2[colnames(norm_exp),]$Condition))], cex=0.7, adj =-0.1) -->
<!-- abline(h=0, v=0) -->
<!-- ``` -->


```{r}
R.avg <- avereps(RG$R, ID=RG$genes$ProbeName)
G.avg <- avereps(RG$G, ID=RG$genes$ProbeName)
colnames(R.avg) = paste(ed[colnames(R.avg),]$Cy5,"_Cy5","_",ed[colnames(R.avg),]$Scan.ID,sep="")
colnames(G.avg) = paste(ed[colnames(G.avg),]$Cy3,"_Cy3","_",ed[colnames(G.avg),]$Scan.ID,sep="")

intensity_matrix = cbind(R.avg, G.avg)
norm_intensity_matrix = normalizeBetweenArrays(intensity_matrix, method="quantile")
```



# Between array normalization using Aquantile

This normalization will only be used in the case of Single Channel Analysis. 

```{r,  fig.width=8, fig.height=8}

# Single Channel analysis - all arrays
MA.n <- normalizeBetweenArrays(MA.avg, method="Aquantile")
plotDensities(MA.n)
```

## Control probes

The following control probes exist on the arrays used in this experiment:

* Corner associated (used for orientation purposes during scanning)
  * Bright corner 
  * Dark corner 
* Negative controls
  * 3xSLv1 (hairpin probe that does not hybridize well with any possible RNA)
* Positive controls
  * Human GAPDH and PGK1 probes
  * Deletion stringency probes (DCP, probe with varying number of insertions/changes with respect to reference; the number after the "_" denotes the number of differences to the reference which should correlate with lower expression)
  * E1A_r60: spike-in probes with concentrations that should cover the whole dynamic range of the array

There are a few other expression probes that are used by Agilent's feature extraction/QC pipeline. 

```{r, qc_probes,  fig.width=12, fig.height=3}
control_probes = which(RG$genes$ControlType!=0)
cp_data = RG$G[control_probes,]
cp_names = RG$genes[control_probes,]
selected_controls = ifelse(substr(cp_names$ProbeName,1,4) %in% c("ERCC", "ETG0","ETG1") | substr(cp_names$ProbeName,1,2) == "RC",F,T)

# control probes
for (i in 1:ncol(cp_data)) {
  boxplot(log2(cp_data[selected_controls,i]) ~ factor(cp_names$ProbeName[selected_controls]),las=2, main=paste("Sample",colnames(cp_data)[i]), outline=F, xlab="")
}
```


```{r}
save(ed, RG, MA, MA.n, exp_matrix, file="../Data/Preprocessed/Stomach_antrum_Lgr4_KO_preprocessed_data.Rdata")
```

