---
title: "Microarray Data Stomach Antrum Epithelia, Lgr4 - GSEA analysis on DGE results - Plots"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, warning=FALSE, message=FALSE}
rm(list=ls())

#suppressMessages(library(xlsx))
library(ggplot2)
library(knitr)
library(scales)
library(fgsea)
library(data.table)

source("./fGSEA_plot_functions.R")

result_folder = "../Results/"

set.seed(12345)
```

# Overview

This is data from a stomach antrum samples of various mouse experiments (Lgr4 KO vs WT) hybridized as dual channel libraries to Agilent 014568 micro arrays. 

For analysis, a preranked analysis using the R package fGSEA was used that should give similar results to preranked analysis in standard GSEA. 


```{r}
load(file.path(result_folder, "DGE_results.Rdata"))
```

```{r}
data_generation_date = "2022-04-25"
output_folder = result_folder
GSEA_datafile = file.path(output_folder,"fGSEA_Results.Rdata")
load(GSEA_datafile)
load(file.path(output_folder, "Geneset_collections.Rdata"))
```


The plots below use differential gene expression data and GSEA results from `r data_generation_date` .


# MSigDB gene sets

Since MSigDB gene sets use human gene symbols to map genes to pathways we translated mouse symbols to homologous human symbols using HomologeneDB from NCBI (build 68). 


```{r}
homologene_human_mouse = fread("S:/AG/AG-Sigal/Research/Documentation/Projects/Hilmar Berger/References/HomologousGenes/HomoloGene/build68/homologene_9606_10090.txt")
all_entrez_ids = data.frame(EntrezID=unique(all_results[[1]]$EntrezID))
all_symbols = data.frame(GeneSymbol=unique(all_results[[1]]$GeneSymbol))
a1 = merge(all_symbols, homologene_human_mouse[,c("entrez_1","symbol_1","symbol_2"), with=F], by.x="GeneSymbol", by.y="symbol_2", all.x=T, sort=F)

mouse_to_human=a1
mouse_to_human = subset(mouse_to_human, !is.na(GeneSymbol) & !is.na(symbol_1) & !symbol_1 == "")
rownames(mouse_to_human) = mouse_to_human$GeneSymbol
```


# Plots for selected gene sets

## Comparison Lgr4 KO vs. Lgr4 +/+, Hp. infection


```{r}
selected_genesets = c("Lgr5_SC_Signature_Munoz", "Barker_Antral_SC", "Herbst_et_al_Beta_catenin_targets","HALLMARK_TNFA_SIGNALING_VIA_NFKB")
```


```{r}
sel_comparison = "Lgr4_KO_vs_control"
```

### Mouse gene sets

```{r}
tmp = all_results[[sel_comparison]]
ranks = unlist(tapply(tmp$t, tmp$GeneSymbol, function(x) x[which(abs(x)==max(abs(x)))] ))

for (g in selected_genesets) {
  rr = subset(all_gsea_results[[sel_comparison]], pathway==g)
  if(nrow(rr)==0) warning(paste0("Gene set ", g, "not found"))
  loc = unlist(lapply(gset_collections_mouse, function(x) grep(g, names(x))))
  if(length(loc) == 0) next 
  sel_genes = unlist(gset_collections_mouse[[names(loc)]][loc])
  p = myPlotEnrichment(sel_genes, ranks)
  print(p + ggtitle(g, paste("FDR=",prettyNum(rr$FDR_global, digits=2),", p=",prettyNum(rr$pval, digits=2), ", ES=",prettyNum(rr$ES, digits=2), ", NES=",prettyNum(rr$NES, digits=2) ) ) )
}
```


```{r}
for (g in selected_genesets) {
  rr = subset(all_gsea_results[[sel_comparison]], pathway==g)
  if(nrow(rr)==0) warning(paste0("Gene set ", g, "not found"))
  loc = unlist(lapply(gset_collections_mouse, function(x) grep(g, names(x))))
  if(length(loc) == 0) next
  sel_genes = unlist(gset_collections_mouse[[names(loc)]][loc])
  p = myPlotEnrichment_pdf(sel_genes, ranks, linewidth = 2)
  pdf(file=file.path(result_folder, paste0(sel_comparison,"_",g, ".pdf")), width=8, height = 6)
  print(p + ggtitle(g, paste("FDR=",prettyNum(rr$FDR_global, digits=2),", p=",prettyNum(rr$pval, digits=2), ", ES=",prettyNum(rr$ES, digits=2), ", NES=",prettyNum(rr$NES, digits=2) ) ) )
  dev.off()
}
```


```{r}
tmp = all_results[[sel_comparison]]
tmp2 = merge(tmp, mouse_to_human, by="GeneSymbol")
tmp2 = subset(tmp2, !is.na(symbol_1))
ranks = unlist(tapply(tmp2$t, tmp2$symbol_1, function(x) x[which(abs(x)==max(abs(x)))] ))

for (g in selected_genesets) {
  rr = subset(all_gsea_results[[sel_comparison]], pathway==g)
  if(nrow(rr)==0) warning(paste0("Gene set ", g, "not found"))
  loc = unlist(lapply(gset_collections_human, function(x) grep(g, names(x))))
  if(length(loc) == 0) next 
  sel_genes = unlist(gset_collections_human[[names(loc)]][loc])
  p = myPlotEnrichment(sel_genes, ranks)
  print(p + ggtitle(g, paste("FDR=",prettyNum(rr$FDR_global, digits=2),", p=",prettyNum(rr$pval, digits=2), ", ES=",prettyNum(rr$ES, digits=2), ", NES=",prettyNum(rr$NES, digits=2) ) ) )
}
```

```{r}
for (g in selected_genesets) {
  rr = subset(all_gsea_results[[sel_comparison]], pathway==g)
  loc = unlist(lapply(gset_collections_human, function(x) grep(g, names(x))))
  if(length(loc) == 0) next 
  sel_genes = unlist(gset_collections_human[[names(loc)]][loc])
  p = myPlotEnrichment_pdf(sel_genes, ranks, linewidth = 2)
  pdf(file=file.path(result_folder, paste0(sel_comparison,"_",g, ".pdf")), width=8, height = 6)
  print(p + ggtitle(g, paste("FDR=",prettyNum(rr$FDR_global, digits=2),", p=",prettyNum(rr$pval, digits=2), ", ES=",prettyNum(rr$ES, digits=2), ", NES=",prettyNum(rr$NES, digits=2) ) ) )
  dev.off()
}
```


### PDF export

All plots have been exported to corresponding files in PDF format in folder `r result_folder`.


# Software versions

```{r}
sessionInfo()
```
